# Introduction

The goal of this analysis is to see how will phylogenetically independent 
contrasts can reconstruct character covariance when the number of 
characters exceeds the number of species. In this case we will look at 30 
characters on a tree with 8 species.

# Implementation

Get set up, and construct the covariance matrix

```{r}
library(geiger, verbose=FALSE)
library(Matrix)
library(picante)

TrueCov = bdiag(matrix(rep(0.9, 100), 10), matrix(rep(0.5, 100), 10), diag(10))
diag(TrueCov) = 1
TrueCov = as.matrix(TrueCov)
```

Simulate the tree and data
```{r fig.width=7, fig.height=6}
tree = rcoal(8)
plot(tree)
D = sim.char(tree, TrueCov, nsim=1, model='BM')
```

Calculate the independent contrasts
```{r}
contrasts = apply( D, 2, function(a) pic(a, tree) )
```

Calculate the observed correlations from the contrasts
```{r}
ObsCor = cor.table(contrasts)$r
```
Display the original and observed matrices

```{r graphics}
image(TrueCov, main='Original')
image(ObsCor, main='Observed')
```

# Interpretation

Relative to the true matrix, the observed matrix is quite noisy. Many of 
the elements that should be 0 are instead estimated to be quite strong, 
and there is quite a bit of variation within blocks that should be 
non-zero.

This means that we have to be careful interpretting covariances in 
comparative phylogenetic analyses when the number of variables exceeds 
the number of species. In particular, there can be many false positives.

# Building this document

To compile an html report from source, execute the following in R.

```{r compile, eval=FALSE}
library(knitr)
knit2html('my_analysis.Rmd')
```

The following provides details on the versions of tools that were used to 
generate the report.

```{r}
sessionInfo()
```